Paper 4: quasi-experiment

### 2024-10-08 Rmd/Paper4.Rmd
Prepared for publication

---
title: "Paper 4"
author: "Niklas Karlsen"
date: "2024-08-22"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), '../output/paper4.html')) })
---

This is a computational essay?

Changelog

* 22.8.2024: added comparison of two hybrid models

TODO

# Preamble / technical details

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #,results="asis")

Sys.setenv(LANG = "no_NB.UTF-8")            # for æ,ø og å. Kilde:
Sys.setlocale("LC_ALL","Norwegian")
# https://stackoverflow.com/questions/16347731/how-to-change-the-locale-of-r

# variable
plot_to_file = 0
```

```{r last_bibliotek, echo=FALSE, warning=FALSE, message=FALSE}
## Last bibliotek: rekkefølge - de siste overstyrer de første

## bibliotek for tabell-behandling
# last plyr før dplyr
#library(plyr)                               # for plot of mean and sd
library(dplyr)
library(tidyr)
library(pander)                             # prettyprint for tables
# kilde: https://stackoverflow.com/questions/31626419/in-knitr-no-output-from-pander-in-for-loop
#panderOptions('knitr.auto.asis', FALSE)     # to pander in loop
#library(tidyverse)

## bibliotek for plotting
#library(plotrix)                            # SE(m) (plot mean+CI)
library(ggplot2)
#library(GGally)                             # for plot av ggpairs
# kilde: https://ggplot2-book.org/arranging-plots.html#arranging-plots
#library(patchwork)                          # for inset_element

## bibliotek for statistikk
#library(bruceR,quietly=TRUE)                           # for manova
#library(DescTools)                          # for lambda, categorical data
library(psych)                              # for ICC, SD
library(effsize)                            # for cohen's d
#library(corrplot)
#library(rcompanion)                          # for Cramers V
#library(Rfit)                               # ranked regression # https://journal.r-project.org/archive/2013/RJ-2013-027/RJ-2013-027.pdf

### regression libraries
library(lme4)                                 # for MLM
#library(lmerTest)                             # not recommended? Avoid the lmerTest package (https://utstat.toronto.edu/~brunner/workshops/mixed/NormalWithR.pdf)
library(car)                                   # # for p-values i lme4 (anova, test="F"). durbinWatson
library(nlme)                                 # for lme
#library(rmcorr)
library(plm)                                  # paneldata analysis

@@

# SE calculation
library(pcse)                                 # panel-corrected SE
library(sandwich) # cluster corrected error
library(lmtest) # logLikelihood tests, AIC, BIC ,...
#library(MuMIn) # AICc
library(aod) # wald.test

library(texreg)                               # screenreg (print compare results regresion)
library(viridis)

#library(ppcor) # partiell correlation
#library(mice) # missing
library(MuMIn) # AICc

@@

## Definer noen hjelpefunksjoner.

```{r last_hjelpefunksjoner}
source("../../scripts/help-functions.R")
```

@@

```{r load_constructs}
source("../../scripts/constructs.R")
```

The purpose of this script is to do the calculations presented in paper 4 for prototype 1 and 2.

@@

Properties of samples

@@

add group information

```{r add_groups_v22}
# add column with group numbers
data_emne2_v22$group <- c(1:nrow(data_emne2_v22))
data_emne2_v22[data_emne2_v22$Subject == 'Gabriel' |
                 data_emne2_v22$Subject == 'Hulda' |
                 data_emne2_v22$Subject == 'Isak' |
                 data_emne2_v22$Subject == 'Maria' |
                 data_emne2_v22$Subject == 'Nicoline',"group"] <-
                 "armageddon" #1

data_emne2_v22[data_emne2_v22$Subject == 'Conny' |
                 data_emne2_v22$Subject == 'Fride' |
                 data_emne2_v22$Subject == 'Jonas' |
                 data_emne2_v22$Subject == 'Julia' |
                 data_emne2_v22$Subject == 'Lea',"group"] <-
                 "fasetilstander" # 2

data_emne2_v22[data_emne2_v22$Subject == 'Berta' |
                 data_emne2_v22$Subject == 'Dina' |
                 data_emne2_v22$Subject == 'Gloria' |
                 data_emne2_v22$Subject == 'Klaudia',"group"] <- 
                 "atomeksitasjon" # 3

data_emne2_v22[data_emne2_v22$Subject == 'Benjamin' |
                 data_emne2_v22$Subject == 'Ester' |
                 data_emne2_v22$Subject == 'Harald' |
                 data_emne2_v22$Subject == 'Inga' |
                 data_emne2_v22$Subject == 'Olava',"group"] <- 
                 "smittespredning" # 4

data_emne2_v22[data_emne2_v22$Subject == 'David' |
                 data_emne2_v22$Subject == 'Elias' |
                 data_emne2_v22$Subject == 'Filip',"group"] <-
                 "trykkbeholder" # 5

data_emne2_v22 <- data_emne2_v22 %>% relocate(group, .after = Subject)           # bruker dplyr
```

```{r add_groups_v23}
# add column with group numbers
data_emne2_v23$group <- rep(NA,nrow(data_emne2_v23))
data_emne2_v23[data_emne2_v23$Subject == 'Berta' |
                 data_emne2_v23$Subject == 'Fride' |
                 data_emne2_v23$Subject == 'Gloria' |
                 data_emne2_v23$Subject == 'Nora',"group"] <-
                 "kokeplate" #1/A

data_emne2_v23[data_emne2_v23$Subject == 'Conny' |
                 data_emne2_v23$Subject == 'Idun' |
                 data_emne2_v23$Subject == 'Magrit' |
                 data_emne2_v23$Subject == 'Vaarin',"group"] <-
                 "arv øyefarge" # 2/B

data_emne2_v23[data_emne2_v23$Subject == 'Julia' |
                 data_emne2_v23$Subject == 'Dina' |
                 data_emne2_v23$Subject == 'Ruben',"group"] <- 
                 "partikkelmodellen gass" # 3/C

data_emne2_v23[data_emne2_v23$Subject == 'Emmanuel' |
                 data_emne2_v23$Subject == 'Klaudia' |
                 data_emne2_v23$Subject == 'Noah' | 
                 data_emne2_v23$Subject == "Christian","group"] <- 
                 "partikkelmodellen vannglass" # 4/E

data_emne2_v23[data_emne2_v23$Subject == 'Gabriel' |
                 data_emne2_v23$Subject == 'Harald' |
                 data_emne2_v23$Subject == 'Lukas' |
                 data_emne2_v23$Subject == 'Mikael',"group"] <-
                 "friksjon" # 5/F

data_emne2_v23[data_emne2_v23$Subject == 'Olava' |
               data_emne2_v23$Subject == "Yosita" | 
               data_emne2_v23$Subject == "Nicoline" ,"group"] <-
                 "faseoverganger grønn rute" # 6/D

data_emne2_v23 <- data_emne2_v23 %>% relocate(group, .after = Subject)           # bruker dplyr
```

@@

Data were collected using the measure developed by Karlsen et al. (2024) / paper 1.

@@

@@

To decide which statistical tests we should use to compare pre-post scores, the Levene’s (Mean) test for equal variances and the Kolmogorov-Smirnov test for normality were used as preliminary tests (Pearce and Derrick, 2019). Due to the small sample sizes, the data are probably non-normally distributed. If the variance of the pre-post differences are unequal, this suggests the use of the non-parametric Yuen d-test with trimmed means (Mair & Wilcox, 2020), which show good power for low samples (Wanishsakpong et al., 2023) and can be appropriate when one is not interested in generalizing to a population (Ng  & Cribbie, 2019). If, on the other hand, the pre-post data have equal variance, this suggests the use of the Mann-Whitney (i.e., Wilcox) test with arithmetic mean.

@@

beregn SID (Estrada et al., 2019).

@@

Ref.: Marsden og Torgerson (2012)

@@

@@

Hybrid with variable coefficients was chosen since there was only one significant dependent variable in the fixed-effects model (Bell et al., 2019).

@@

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TK)
```

@@

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TK)
```

@@

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22)

summary(hybrid_TK)
```

@@

Use FIML (ML) for model comparison, REML for parameter estimates (Christophersen, 2018).

@@

summary of details for best fit models

@@

>>>>

---
title: "Paper 4: quasi-experiment"
author: "Niklas Karlsen, OsloMet, Norway, e-mail: niklaska@oslomet.no"
date: "2024-10-08"
bibliography: "../data/meta_data/references.bib"
csl:          "../scripts/university-of-gothenburg-apa-7th-edition-swedish-legislations.csl" # (source: https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html)
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), '../output/Paper4.html')) })
---

The purpose of this script is to present the results of the quasi-experiment in paper 4, by presenting the calculations presented in paper 4 for prototype 1 and 2. Three effect sizes are calculated:

## Load libraries

```{r last_bibliotek}
## bibliotek for tabell-behandling
library(dplyr)
library(tidyr)
library(pander)                             # prettyprint for tables
library(texreg)                               # screenreg (print compare results regresion)

## bibliotek for plotting
library(ggplot2)
library(viridis)

## bibliotek for statistikk
library(psych)                              # for ICC, SD
library(effsize)                            # for cohen's d

### regression libraries
library(lme4)                                 # for MLM
#library(lmerTest)                             # not recommended? Avoid the lmerTest package (https://utstat.toronto.edu/~brunner/workshops/mixed/NormalWithR.pdf)
library(car)                                   # # for p-values i lme4 (anova, test="F"). durbinWatson
library(nlme)                                 # for lme
library(plm)                                  # paneldata analysis

@@

# SE calculation
library(pcse)                                 # panel-corrected SE
library(sandwich) # cluster corrected error
library(lmtest) # logLikelihood tests, AIC, BIC ,...
library(aod) # wald.test
library(MuMIn) # AICc

@@

@@

These are the items that were used. The TK-items differ somewhat from the items used in @karlsen_assessing_2024.

```{r load_constructs}
#source("../../scripts/constructs.R")

# TPACK-PST 22
TK_PST22_pos <- c("TK2","TK5","TK6") # ,"TK5","TK6","TCK1"
PCK_PST22_pos <- c("CK1","CK2","CK3","CK5","PCK2","PCK5") # "CK2", "CK5",
TPACK_PST22_pos <- c("TCK1","TCK4","TPK1","TPK4","TPACK2", "TPACK4")

TPACK_PST_emne2_v22 <- list("TK"=TK_PST22_pos,
                            "PCK"=PCK_PST22_pos,
                            "TPACK"=TPACK_PST22_pos)

# TPACK-PST 22 adjusted to 23
TK_PST23_pos <- c("TK2","TK5","TK6") # "TK4","TK5","TK6"
PCK_PST23_pos <- c("CK123", "PCK2")
TPACK_PST23_pos <- c("TPACK2","TCK4","TPK1","TPK4","TPACK4") 
TPACK_PST_emne2_v23 <- list("TK"=TK_PST23_pos,
                            "PCK"=PCK_PST23_pos, 
                            "TPACK"=TPACK_PST23_pos) #, 
```

@@

### Prepare dataset

@@

@@

Data were collected using the measure developed by @karlsen_assessing_2024 (paper 1).

@@

### Choose subsample of PSTs

@@

To decide which statistical tests we should use to compare pre-post scores, the Levene’s (Mean) test for equal variances and the Kolmogorov-Smirnov test for normality were used as preliminary tests [@pearce_preliminary_2019]. Due to the small sample sizes, the data are probably non-normally distributed. If the variance of the pre-post differences are unequal, this suggests the use of the non-parametric Yuen d-test with trimmed means [@mair_robust_2020], which show good power for low samples [@wanishsakpong_comparison_2023] and can be appropriate when one is not interested in generalizing to a population [@ng_gamma_2019]. If, on the other hand, the pre-post data have equal variance, this suggests the use of the Mann-Whitney (i.e., Wilcox) test with arithmetic mean.

@@

beregn SID [@estrada_statistics_2019].

@@

Ref.: @marsden_single_2012

@@

Fixed effects were used, comparing only pre-post.

@@

Hybrid with variable coefficients was chosen since there was only one significant dependent variable in the fixed-effects model [@bell_fixed_2019].

@@

hybrid_TPACK_n19 <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TPACK_n19)
```

@@

hybrid_TPACK_varcof <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TPACK_varcof)
```

@@

hybrid_TPACK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22)

summary(hybrid_TPACK)
```

@@

Use FIML (ML) for model comparison, REML for parameter estimates [@christophersen_introduksjon_2018].

@@

#### Summary of details for best fit models

@@

Here I reproduce the results presented in paper 4.

## Difference

```{r}
# v22
pre_scores_v22 <- describe(scores_v22_pre$scores)
post_scores_v22 <- describe(scores_v22_post$scores)

diff_TK_v22 <- yuend(scores_v22_post$scores[,"TK"],
                  scores_v22_pre$scores[,"TK"]
                  )
diff_PCK_v22 <- yuend(scores_v22_post$scores[,"PCK"],
                  scores_v22_pre$scores[,"PCK"]
                  )
diff_TPACK_v22 <- yuend(scores_v22_post$scores[,"TPACK"],
                  scores_v22_pre$scores[,"TPACK"]
                  )

# v23
pre_scores_v23 <- describe(scores_v23_pre$scores)
post_scores_v23 <- describe(scores_v23_post$scores)

diff_TK_v23 <- post_scores_v23["TK","mean"] - pre_scores_v23["TK","mean"]
diff_p_TK_v23 <- wilcox.test(scores_v23_post$scores[,"TK"], scores_v23_pre$scores[,"TK"], paired=T)$p.value

diff_PCK_v23 <- post_scores_v23["PCK","mean"] - pre_scores_v23["PCK","mean"]
diff_p_PCK_v23 <- wilcox.test(scores_v23_post$scores[,"PCK"], scores_v23_pre$scores[,"PCK"], paired=T)$p.value

diff_TPACK_v23 <- post_scores_v23["TPACK","mean"] - pre_scores_v23["TPACK","mean"]
diff_p_TPACK_v23 <- wilcox.test(scores_v23_post$scores[,"TPACK"], scores_v23_pre$scores[,"TPACK"], paired=T)$p.value

# create table
rad_TK_v22 <- list("TK",
  pre_scores_v22["TK","mean"],
            pre_scores_v22["TK","sd"],
            post_scores_v22["TK","mean"],
            post_scores_v22["TK","sd"],
            paste0(format(diff_TK_v22$diff,digit=2),
                   " (p = ", 
                   format(diff_TK_v22$p.value, digits=2), 
                   ")"))

rad_PCK_v22 <- list("PCK",
  pre_scores_v22["PCK","mean"],
            pre_scores_v22["PCK","sd"],
            post_scores_v22["PCK","mean"],
            post_scores_v22["PCK","sd"],
            paste0(format(diff_PCK_v22$diff,digit=2),
                   " (p = ", 
                   format(diff_PCK_v22$p.value, digits=2), 
                   ")"))

rad_TPACK_v22 <- list("TPACK",
  pre_scores_v22["TPACK","mean"],
            pre_scores_v22["TPACK","sd"],
            post_scores_v22["TPACK","mean"],
            post_scores_v22["TPACK","sd"],
            paste0(format(diff_TPACK_v22$diff,digit=2),
                   " (p = ", 
                   format(diff_TPACK_v22$p.value, digits=2), 
                   ")"))

rad_TK_v23 <- list("TK",
            pre_scores_v23["TK","mean"],
            pre_scores_v23["TK","sd"],
            post_scores_v23["TK","mean"],
            post_scores_v23["TK","sd"],
            paste0(format(diff_TK_v23, digit=2),
                   " (p = ",
                   format(diff_p_TK_v23, digit=2),
                   ")")
            )

rad_PCK_v23 <- list("PCK",
            pre_scores_v23["PCK","mean"],
            pre_scores_v23["PCK","sd"],
            post_scores_v23["PCK","mean"],
            post_scores_v23["PCK","sd"],
            paste0(format(diff_PCK_v23, digit=2),
                   " (p = ",
                   format(diff_p_PCK_v23, digit=2),
                   ")")
            )

rad_TPACK_v23 <- list("TPACK",
            pre_scores_v23["TPACK","mean"],
            pre_scores_v23["TPACK","sd"],
            post_scores_v23["TPACK","mean"],
            post_scores_v23["TPACK","sd"],
            paste0(format(diff_TPACK_v23, digit=2, nsmall=2),
                   " (p = ",
                   format(diff_p_TPACK_v23, digit=2, nsmall=2),
                   ")")
            )

row1_p1 <- c("", "Prototype 1", "", "", "", "")
row1_p2 <- c("", "Prototype 2", "", "", "", "")
row2 <- c("", "Pre", "", "Post", "", "")
row3 <- c("", "Mean", "SD", "Mean", "SD", "diff")

prepost_table_v22 <- t(data.frame(
  row1_p1, row2, row3,
  format(rad_TK_v22,digit=2,nsmall=2), # only 2 digits
  format(rad_PCK_v22,digit=2,nsmall=2), # only 2 digits
  format(rad_TPACK_v22,digit=2,nsmall=2) # only 2 digits
))

prepost_table_v23 <- t(data.frame(
  row1_p2, row2, row3,
  format(rad_TK_v23,digit=2,nsmall=2), # only 2 digits
  format(rad_PCK_v23,digit=2,nsmall=2), # only 2 digits
  format(rad_TPACK_v23,digit=2,nsmall=2) # only 2 digits
))

rownames(prepost_table_v22) <- NULL # remove row names
rownames(prepost_table_v23) <- NULL # remove row names

pander(prepost_table_v22)
pander(prepost_table_v23)
```

## Fixed effects

```{r}
screenreg(list("within22" = within22,
               "within23pp" = within23_prepost))
```

## Within-between

```{r}
summary(hybrid_TPACK)
```

TK_S and PCK_S are the within-effects, while TK_mean and PCK_mean are the between-effects.

This model is however not directly comparably to the fixed effects model, since the hybrid model uses all the students in the analysis (n=21+20), while the fixed effects only uses the students that are in both pre-post datasets (n=19+19) (cf. model hybrid_TPACK_n19 above).

```{r}
# print summary of models
screenreg(list("Hybrid (all students)" = hybrid_TPACK, 
               "Hybrid (pre-post students)" = hybrid_TPACK_n19))
```

# Discussion

The hybrid model reported in paper 4, should perhaps be switched to the n=19 model, for more direct comparison with the fixed effects model.

# References

<!-- citation(package) -->

<div id="refs"></div>

<!-- source: https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html

# (APPENDIX) Appendix {-} -->

# Session info
<!-- # ref.: https://intro2r.com/proj_doc.html -->
<!-- #sessionInfo() -->

```{r}
xfun::session_info()
```
